\large{\bf Quantitative genetics with stable laws }

### 2023-06-11 writeup.tex
histograms of results

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage{natbib}
\usepackage{xcolor}

@@

It is mathematically intriguing to speculate what effect a non-Gaussian distribution
for $\epsilon$ would have on the dynamics induced by~\eqref{eqn:offspring_model}.
But, is there any evidence that non-Gaussian segregation noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 individuals from the March 2018 release of the UK Biobank.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of SNPs on the phenotype relative to the reference allele.
The phenotypes we consider were all coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(so, effects are \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Suppose that the phenotype of each individual is additive:
the maternal trait value is $x = \sum_i x_{i,1} + x_{i,2}$
and the paternal trait value is $y = \sum_i y_{i,1} + y_{i,2}$,
where $x_{i,1}$ and $x_{i,2}$ are the effects of the alleles carried
by the mother on each of the two chromosomes
at the $i^\text{th}$ locus, respectively.
Then the offspring's trait is $z = \sum_i z_{i,m} + z_{i,p}$,
where $z_{i,m} = \theta_{i,m} x_{i,1} + (1-\theta_{i,m}) x_{i,2}$,
and $\theta_{i,m} = 1$ with probability 1/2 and $\theta_{i,m} = 0$ otherwise;
and likewise $z_{i,p} = \theta_{i,p} y_{i,1} + (1-\theta_{i,p}) y_{i,2}$.
With this notation, the segregation noise $z - \frac{x+y}{2}$
is a sum of terms of the form
$$
    \left( \theta_{i,m} - 1/2 \right) x_{i,1}
    + \left( 1/2 - \theta_{i,m} \right) x_{i,2} 
$$
(and similar terms with $y$ in place of $x$).
Now let $X$ be the effect of a uniformly chosen allele carried by a uniformly chosen individual
at a uniformly chosen locus.
The appropriate central limit theorem is determined by the tail behavior of $|(\theta - 1/2) X| = |X|/2$;
i.e., we'd like to know
$$
    \lim_{t \to \infty} \frac{ \log \P\{ |X| > 2 t \} }{ \log t } = \alpha ,
$$
if the limit exists.
Let $p_i$ be the minor allele frequency at the $i^\text{th}$ locus;
since the data we have assigns effect 0 to the major allele,
$x_i = 0$ with probability $1-p_i$.
Therefore, if we have estimated effects for $N$ SNPs
with minor allele frequency $p_i$ and effect size $e_i$ for the minor allele,
for $1 \le i \le N$,
then
$$
    \P\{ |X| > 2t \} \approx \frac{1}{N} \sum_{i : |e_i| > 2t} p_i .
$$
So, to obtain a per-trait estimate of $\alpha$,
we plot $\frac{1}{N} \sum_{i : |e_i| > t} p_i$ against $t$
for a sequence of values of $t$,
and fit a linear model to the tail
(which we take to be all values of $t$ for which the number of SNPs with absolute effect above $t$
is between 20 and 10\% of all SNPs).
Examples are shown in figure \comment{XXX}.

@@

>>>>

\documentclass{article}
\usepackage{fullpage}
\usepackage{amsmath, amssymb}
\usepackage[hidelinks]{hyperref}
\usepackage{natbib}
\usepackage{xcolor}
\usepackage{graphics}

@@

It is mathematically intriguing to speculate what effect a non-Gaussian distribution
for $\epsilon$ would have on the dynamics induced by~\eqref{eqn:offspring_model}.
But, is there any evidence that non-Gaussian segregation noise is important in practice?
In this paper we focus primarily on the theory,
but first take a brief and imperfect look at what some data have to say on the topic.
To do this, we downloaded the GWAS results provided by~\citet{biobankSNPs}
for 7,221 phenotypes from 361,194 individuals from the March 2018 release of the UK Biobank.
After filtering (see details below) we were left with 695 binary illness-related phenotypes.
Taken at face value, these provide for each phenotype a set of estimates of the additive effect
of the alternative allele at a set of SNPs on the phenotype relative to the reference allele.
The phenotypes we consider were all binary and coded as 0/1,
so estimates could be thought of as additive adjustments to the risk
(i.e., effects are \emph{not} on a logit scale).
These estimates were obtained as the coefficient for the SNP in a simple linear model
fit to each phenotype for each SNP,
with the first 20 principal components and all combinations of age, age squared, and inferred sex as covariates;
we considered effects of those SNPs with a reported $p$-value less than $10^{-8}$.
There are a great many possible issues with these data;
however, the results seem consistent across many SNPs and phenotypes,
and we know of no reason that possible methodological artifacts
(e.g., confounding by uncorrected variables)
would specifically affect the decay of the tail of effect sizes.

Suppose that the phenotype of each individual is additive:
the maternal trait value is $x = \sum_i x_{i,1} + x_{i,2}$
and the paternal trait value is $y = \sum_i y_{i,1} + y_{i,2}$,
where $x_{i,1}$ and $x_{i,2}$ are the effects of the alleles carried
by the mother on each of the two chromosomes
at the $i^\text{th}$ locus, respectively.
Then the offspring's trait is $z = \sum_i z_{i,m} + z_{i,p}$,
where $z_{i,m} = \theta_{i,m} x_{i,1} + (1-\theta_{i,m}) x_{i,2}$,
and $\theta_{i,m} = 1$ with probability 1/2 and $\theta_{i,m} = 0$ otherwise;
and likewise $z_{i,p} = \theta_{i,p} y_{i,1} + (1-\theta_{i,p}) y_{i,2}$.
With this notation, the segregation noise $z - \frac{x+y}{2}$
is a sum of terms of the form
$$
    \left( \theta_{i,m} - 1/2 \right) x_{i,1}
    + \left( 1/2 - \theta_{i,m} \right) x_{i,2} 
$$
(and similar terms with $y$ in place of $x$).
Now let $X$ be the effect of a uniformly chosen allele carried by a uniformly chosen individual
at a uniformly chosen locus.
The appropriate central limit theorem is determined by the tail behavior of $|(\theta - 1/2) X| = |X|/2$.
So, we'd like to know the value of
$$
    \lim_{t \to \infty} \frac{ \log \P\{ |X| > 2 t \} }{ \log t } = \alpha ,
$$
if the limit exists.
Let $p_i$ be the minor allele frequency at the $i^\text{th}$ locus;
since the data we have assigns effect 0 to the major allele,
$x_i = 0$ with probability $1-p_i$.
Therefore, if we have estimated effects for $N$ SNPs
with minor allele frequency $p_i$ and effect size $e_i$ for the minor allele,
for $1 \le i \le N$,
then
$$
    \P\{ |X| > 2t \} \approx \frac{1}{N} \sum_{i : |e_i| > 2t} p_i .
$$
So, to obtain a per-trait estimate of $\alpha$,
we plot $\log \frac{1}{N} \sum_{i : |e_i| > t} p_i$ against $\log t$
for a sequence of values of $t$,
and fit a linear model to the tail
(which we take to be all values of $t$ for which the number of SNPs with absolute effect above $t$
is between 20 and 10\% of all SNPs).
Examples are shown in figure~\ref{fig:example_snps}.

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/examples}
    \end{center}
    \caption{
        TODO: cut this down to just 2 SNPs and move this figure to the supplement; change 'x' to 't' in axis labels.
        Tail distributions of frequency-weighted effects
        for twelve randomly chosen phenotypes.
        In each, for a phenotype with $N$ SNPs,
        effect sizes $e_i$ and frequencies $p_i$,
        the black line shows
        $\log \frac{1}{N} \sum_{i : |e_i| > t} p_i$ (vertical axis)
        against $\log t$ (horizontal axis);
        the red line shows the least-squares linear fit;
        each plot only covers the 10\% largest-effect SNPs.
        Legends show the coefficients of the red line; the slope on `x' is $\alpha$,
        the estimated exponent;
        ``num snps'' is the number of SNPs for the phenotype with $p$-value less than $10^{-8}$,
        and ``num cases'' is the reported number of the rougly 360,000 individuals
        recorded as having the listed phenotype.
        \label{fig:example_snps}
    }
\end{figure}

We applied this procedure to results from models fit on both sexes,
after removed phenotypes related to employment, drug treatments, and diet.
This left us with 1,108 phenotypes for which which could estimate the tail exponent.
For the main results,
we then restricted to phenotypes having non-missing numbers of ``cases'' and ``controls''
and a sample size of at least 300,000 (i.e., not more than 61,194 of 361,194 missing phenotypes).
We also removed
the 65 remaining phenotypes that are common (more than 10,000 cases),
as these often showed very different patterns
(e.g., effect size distributions).
This left us with 695 phenotypes;
plots including the 413 ``filtered'' phenotypes are provided in
Supplementary Figure~\ref{sfig:unfiltered_hist}.

Resulting values of the tail exponent $\alpha$ are shown in Figure~\ref{fig:exponent_hist}.
Estimated tail exponents range from about 1 to 2.5,
and 28.9\% had estimated values less than 1.5.
A concern could be that ``noisier'' phenotypes,
for which estimated effect sizes are less reliable,
might lead to larger estimated tail exponents;
for this reasons, we looked for an association between
tail exponent and two proxies for power,
number of SNPs (i.e., number of SNPs with $p$-value less than $10^{-8}$)
and number of cases.
Interestingly, the result in nonmonotonic,
with exponents closer to $\alpha=2$ for phenotypes with around 1000 cases,
and lower exponents for rarer \emph{and} more common phenotypes.
(A similar pattern is seen for ``number of SNPs'',
but this may be due to association with number of cases.)
This may indicate a statistical artifact,
or it may be a result of biological differences in genetic architecture
between more and less common phenotypes.

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/results_10}
    \end{center}
    \caption{
        Estimated values of the tail exponent, $\alpha$,
        for 775 illness-related binary phenotypes:
        \textbf{(A)} distribution of values; and plotted against
        \textbf{(B)} number of SNPs, and
        \textbf{(C)} number of cases.
        \label{fig:exponent_hist}
    }
\end{figure}

@@

\appendix

\begin{figure}
    \begin{center}
    \includegraphics{snp_effects/unfiltered_results_10}
    \end{center}
    \caption{
        Estimated values of the tail exponent, $\alpha$,
        for 1108 illness-related binary phenotypes,
        including the 333 phenotypes removed in filtering,
        which are shown as red points in lower plots.
        \textbf{(A)} distribution of values; and plotted against
        \textbf{(B)} number of SNPs,
        \textbf{(C)} number of cases, and
        \textbf{(C)} number of non-missing subjects.
        \label{fig:unfiltered_hist}
    }
\end{figure}
